---
title: "Gene expression relating to macrophages for trauma repair"
subtitle: "Project for statistical assessment of a gene expression file with R studio"
author: "L T Stein"
date: "2/14/2023"
output: pdf_document
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
library(pander)
library(DESeq2)
library(affy)
library(pheatmap)
library(PoiClaClu)
library(ggplot2)
```

\newpage

# Summary

Summary about this project


# Introduction

Introduction about this project

\newpage

# 1. Loading data into R



The gene expression file in .tsv format is loaded with read.table(), since
the data includes tab separated data *sep = '\\t' * is specified. Also a header is 
included so that must be notified with *header = TRUE *. Finally three rows are
printed using pander() and head().

```{r}
count_data <- read.table(paste0("/homes/ltstein/Kwartaal_7/Thema_opdracht/GSE215801_exp_counts.tsv/", 
                        "GSE215801_exp_counts.tsv"), sep = '\t', header = TRUE)
colnames(count_data) <- c("Untreated_1", "IL4-treated_1", "PGE2-treated_1", 
                         "IFNg-treated_1", "Untreated_2", "IL4-treated_2",
                         "PGE2-treated_2", "IFNg-treated_2", "Untreated_3", 
                         "IL4-treated_3", "PGE2-treated_3", "IFNg-treated_3")
```

```{r annotation, echo = FALSE}
pander(head(count_data[1:5], 1))
```
*Table 1.1: Expression data of gene counts from .tsv file* 



In the table above the expression data is displayed of the .tsv file *"GSE215801_exp_counts.tsv" *. \
The x-axis is divided into twelve samples and three batches, test groups. Additionally on the y-axis all the gene names are displayed.

\break

In the code block below dim() prints the dimensions of the dataset and str() the structure.


```{r}
structure <- str(count_data)
cat(structure)
```
```{r}
dimension <- dim(count_data)
cat("dimensions =", dimension)
```

In the first code block above the structure and datatypes are displayed to confirm the data is correct .
Also the second code block the dimensions of the x-as and y-axis are displayed from the dataset.



Case and control is assigned using indexing. This assigning makes a clearer difference between
the experimental group and test group

```{r}
case1 <- c(2,6,10)
case2 <- c(3,7,11)
case3 <- c(4,8,12)
control <- c(1,5,9)
```

In the code above case is assigned to the tested groups and control is assigned to the control group


\newpage
# 2.1 Boxplot

Two boxplots are made of the data using the 'boxplot()' function. Since the raw data of the first boxplot shows a range which is too wide, it doesn't offer a clear view of the data.
In the second boxplot a log-transformation is performed on the data using the 'log2()' function, that narrows down the range.

The replication groups are differentiated using the 'rep()' function.
A vector with three colors is passed to rep, and rep is told that per four items 
the colors switch.


```{r fig.cap= "Boxplot of gene expression of twelve samples"}
boxplot(count_data, las = 2, par(mar=c(8,8,4,2)), main = "Gene expression of twelve samples",
        xlab = "", ylab = "")
mtext("Samples", line = 0.7)
title(ylab = "Gene counts (n)", line = 5)
```


In the figure above a boxplot is performed on the raw data.
On the x-axis lie all the samples and on the y-axis are the gene counts.



```{r fig.cap= "Boxplot of log-transformed gene expression of twelve samples"}
color_place_holder <- rep(c("orange", "cyan", "purple"), each = 4)


boxplot(log2(count_data + 1), las = 2, par(mar=c(8,8,4,2)),
        main = "Gene expression of twelve samples that are log-transformed",
        xlab = "", ylab = "", col= color_place_holder)
mtext("Samples", line = 0.7)
title(ylab = "Gene counts (n)", line = 5)
```


In the figure above a boxplot is performed on the raw data, using a log-transformation.
On the x-axis lie all the samples and on the y-axis are the gene counts.

\newpage

# 2.2 Density plot

The figure below is plotted using the 'plotDensity()' function from the ´affy'
library. The legend is made with the 'legend()' function for all columns
and the colors are specified using the same method as used in the boxplots
above.

Finally a vertical line is added using 'abline()' to indicate that the
values from the peak on its left are all inactive genes.

```{r fig.cap = "Density plot of gene expression data using log-transformation"}
plotDensity(log2(count_data + 0.1), col = color_place_holder,
            main = "Density plot of gene expression data using log-transformation",
            xlab = "log2(count_data)")
legend('topright', names(count_data), lty=c(1:ncol(count_data)),
       col= color_place_holder, cex = 0.95)
abline(v=-1.5, lwd=1, col='red', lty=2)
```


In the figure above a density plot is made from the expression data. On
the right side of the red line are all active genes and on its left side
are therefore all inactive genes. 

The replication group are divided into the same 3 colors as in the boxplots,
and each sample received an individual line.

\newpage

# 3.1 Normalization




```{r fig.cap= "Expression read counts for all samples"}
par(mar=c(7,3,2,2))
barplot((colSums(count_data) / 1e6),
        ylab = "Sequencing depth (times million)",
        xlab  = "",
        main = "Expression read counts for GSE215801",
        col = color_place_holder,
        las = 2)
mtext("Samples", line = -0.5)
legend('topleft', names(count_data), lty=c(1:ncol(count_data)),
       col= color_place_holder, cex = 0.80)
```
In the figure above a barplot is made of the sequencing depth for each sample.
On the x-axis are all samples with divided into three replication groups 
as shown with the colors orange


In the code block below a DESeqDataSet is made with the DESeqDataSetFromMatrix()
function from the 'DESeq2' library

```{r}
(ddsMat <- DESeqDataSetFromMatrix(countData = count_data,
                                  colData = data.frame(samples = names(count_data)),
                                  design = ~ 1))

```

The data is normalized using the *variance stabilizing transformation * method.
In R this method takes shape in the vst() function with the *DESeqDataSetFromMatrix *
as parameter. Later all the normalized values are extracted using the assay() function.

```{r echo = FALSE, results='asis'}
# Normalization step
normalized.data.dds <- vst(ddsMat) 
# Extract values from normalized data object
normalized.data <- assay(normalized.data.dds)

# Tabelize group data
#library(knitr)
#kable(normalized.data[1:4,1:4])
pander(head(normalized.data, 1))

```
*Table 2.1: Normalized data of gene counts from .tsv file* 

In table 2 normalized data is presented for each sample for one gene.

\newpage

# 3.2 Distance calculation

Distances between samples are calculated using the *dist() * function.
The *t (transpose) * function within flips the data to a matrix format.
After calculating all distances between samples that data format is converted
to an actual matrix object using function *as.matrix() *.
Finally the first row is tabelized.
```{r}
# Calculating distances between samples
distance_data <- dist( t ( normalized.data  ))

# convert distance matrix data to an actual matrix object
DistMatrix <- as.matrix(distance_data)

# Tabelize group data
pander(DistMatrix[1:3, 1:4])
```
*Table 2.2: Normalized distance data between samples from .tsv file* 

In table 2.2 is a matrix shown for distances between all samples for one gene.

# 3.3 Sample distances using a Heatmap

```{r fig.cap= "Euclidean distances between samples per group"}
# Get colnames for untreated an treated
Untreated <- c(1,5,9)
Treated <- c(2:4, 6:8, 10:12)
Treatment <- c(1:12)

Treatment[Untreated] <- 0
Treatment[Treated] <- 1
Treatment <- factor(Treatment, labels = c("Untreated", "Treated"))
Treated_genes <- factor(rep(c(0,1,2,3), 3),
                        labels = c("control", "IL4", "PGE2", "IFNg"))

annotation <- data.frame(Treatment, Treated_genes)
annotation                        
rownames(annotation) <- names(count_data)

pheatmap(DistMatrix, show_colnames = FALSE,
         annotation_col = annotation,
         clustering_distance_rows = distance_data,
         clustering_distance_cols = distance_data,
         main = "Euclidean Sample Distances")
```
IFNg-treated 3, IL4-treated 1


\newpage

# 3.4 Multi-Dimensional Scaling

In the code block below several steps are performed for normalizing the raw count data
and calculating preferable distances between the samples called the Poisson Distance.

```{r}
dds <- assay(ddsMat)
poisd <- PoissonDistance( t(dds), type = "deseq")
normalized.samplePoisDistMatrix <- as.matrix(poisd$dd)
mdsPoisData <- data.frame( cmdscale(normalized.samplePoisDistMatrix) )
names(mdsPoisData) <- c('x_coord', 'y_coord')
pander(mdsPoisData)
```

In the figure below a graph is drawn showing the Poisson Distance between
the samples. First a factor is made *factor() * for the three replication groups
assigning unique names to show in the legend. Furthermore the names
of the samples *names() * are assigned to coldata, as the names for 
the samples will show up instead of dots.

The plotting is done with *ggplot() *. The function offers a prettier
and clearer graph and is constructed using more parameters. Such as
the colnames from the raw data, the colors chosen arbitrarily based on the length of groups
and the graph theme is adjusted by *theme_bw *.

```{r fig.cap = "MDS graph showing the Poisson distance between samples"}
groups <- factor(paste(annotation$Treatment, annotation$Treated_genes, sep = "_"))
 
coldata <- names(count_data)


ggplot(mdsPoisData, aes(x_coord, y_coord, color = groups, label = coldata)) + 
  geom_text(size = 4) +
  ggtitle('Multi Dimensional Scaling') +
  labs(x = "Poisson Distance", y = "Poisson Distance") +
  theme_bw()
```
\newpage


# 4.1 Pre-processing

A different way of normalizing the data using the fragments per million mapped fragments (FPM)
formula. As shown below the count data of the samples is divided by the sum of the count data
divided by one million. The log2 is taken from the results plus one, as to remove zero values while
normalization.

```{r}
counts.fpm <- log2( (count_data / (colSums(count_data) / 1e6)) + 1 )
pander(counts.fpm[1:3, 1:5])
```


# 4.1 Assignment

In the code block below genes are kept if they hold a count higher than 
five for at least six samples. If this requirement is not met, then the
gene (or row) will be filtered from the count data. In this case half of
the samples (6) must meet the requirements.

```{r}
# function that receives three parameters, the count_data object,
# a minimum count for a sample as threshold
# and n_samples for amount of samples that satisfy the threshold
normalize_data <- function(count_data, min_count, n_samples){
  keep <- rowSums(count_data >= min_count) >= n_samples
  kept_counts <- nrow(dds[keep,])
  return(kept_counts)
}

filtered_data <- normalize_data(count_data, 5, 6)

original_gene_count <- nrow(count_data)
difference_filtered_count <- filtered_data - original_gene_count
cat("After the filtering step with minimum count 5 for at least 6 samples the new
filtered count is substracted from the original count:", original_gene_count, "-",
filtered_data, "=", difference_filtered_count, "genes")
```

# 4.2 Fold Change

In the code block below the means are calculated for two groups 
(experiment and control) per gene using the *rowMeans() * function.
Then the log fold change is calculated by substracting the control averagescompanying article showing differences in performance compared to other methods and packages.
from the experiments averages. 

Finally a histogram is made with breaks of forty for clearer visualisation of
the deviation. Also an two vertical lines are added at -1 and 1 indicating
low fold-change values in between the borders and vice versa.

```{r fig.cap = "Log fold change of group PGE2-treated"}
# step 1
counts.fpm$experiment_1_avg <- rowMeans(counts.fpm[case2])
counts.fpm$control_avg <- rowMeans(counts.fpm[control])
LFC_column <- counts.fpm$experiment_1_avg - counts.fpm$control_avg
hist(LFC_column, breaks = 40, main = "Log fold change of group PGE2-treated")
abline(v = c(-1 ,1), lt = 2,  lwd = 2, col = "red")
```

In the figure 1.8 above a histogram is made of the log fold change values
between two groups. A clear normal deviation is visible.


# 4.3 Using Bioconductor Packages

In the research's article it's mentioned that the t-test and chi-test are
used for calculating the p-values.

# 4.3.1 The Design (matrix)

Here a matrix model is formed that can be used as the design in the ddsmatrix.

```{r}
Treated_genes
design.matrix <- model.matrix(~ Treated_genes, data=annotation)
design.matrix
```
# 4.3.2 DESeq2


Below a new ddsMat is set up with as main difference passing an factor
with annotated data to 'colData' and a model.matrix is passed as design.

After setting up this matrix passed as an DESeqDataSet  object and
the names of the individual groups are extracted using the 'resultsNames()' function.

Finally these names can be passed to the 'results()' and the summary of each
result object shows the amount of up or down regulated genes for that group,
also an adjusted p-value, outliers, low counts and a mean count.

```{r}
(ddsMat <- DESeqDataSetFromMatrix(countData = count_data,
                                  colData = annotation,
                                  design = design.matrix))

dds <- DESeq(ddsMat)
resultsNames(dds) # lists the names of individual effects
untreated_lfc <- summary(results(dds, name="Intercept"))
Treated_genesIL4_lfc <- summary(results(dds, name="Treated_genesIL4"))
Treated_genesPGE2_lfc <- summary(results(dds, name="Treated_genesPGE2"))
Treated_genesIFNg_lfc <- summary(results(dds, name="Treated_genesIFNg"))
```

A way for removing noise of log2 fold change data of low count genes is by
getting the results from the 'lfcShrink()' instead of the 'results()' function.

It's beneficial to do this since when plotting so that the log fold changes stand out.

```{r}
Treated_genesIL4_shrink <- lfcShrink(dds, coef="Treated_genesIL4", type="apeglm")
Treated_genesPGE2_shrink <- lfcShrink(dds, coef="Treated_genesPGE2", type="apeglm")
Treated_genesIFNg_shrink <- lfcShrink(dds, coef="Treated_genesIFNg", type="apeglm")
```




```{r fig.cap = c("Log fold change of group PGE2-treated", "Log fold change of group PGE2-treated", "Log fold change of group PGE2-treated")}
par(mfrow = c(1,3))
plotMA(Treated_genesIL4_shrink, main = "Log2 fold change of IL4 with shrinkage")
plotMA(Treated_genesPGE2_shrink, main = "Log2 fold change of PGE2 with shrinkage")
plotMA(Treated_genesIFNg_shrink, ylim=c(-6,6), main = "Log2 fold change of IFNg with shrinkage")
```

\newpage

# 5. Data analysis and visualisation

# 5.1 Volcano Plot


\newpage
# Bibliography

 1. GEO Browser. Geraadpleegd op 14-02-2023. De link: https://www.ncbi.nlm.nih.gov/geo/browse/.

 2. Geo Browser. Blood milieu in acute myocardial infarction reprograms human macrophages for trauma repair. Geraadpleegd op 14-02-2023. 
 De link: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE215801
